tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2024-03-12 paper.tex

ARG-based methods can simulate genetic variation for millions of
samples and store the output very compactly
in the ``succinct tree sequence'' encoding~\citep{wong2023general}
and \texttt{tskit} library~\citep{ralph2020}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in gzip-compressed
VCF~\citep{danecek2011}.
At 1.36GB the original simulated ARG
(compressed using the \texttt{tszip} utility) is around 400X smaller.
Furthermore, many calculations can
be expressed efficiently in terms of the underlying
ARG~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.
Finally, outputting a simulated ARG provides access to the full
history, not just the genetic variation among the samples.

@@

\subsection{Implementation and validation}
\texttt{Tstrait} is written entirely in Python. Numerical operations
are either peformed using standard array-oriented
approaches~\citep{numpy} or accelerated
using the \texttt{numba} JIT compiler~\citep{numba}.
The \texttt{tstrait} codebase includes an extensive suite of unit tests,
which are automatically run as part of the development process. The
output of \texttt{tstrait} has been extensively validated against
theoretical expectations, as well as the output
of \texttt{AlphaSimR}~\citep{gaynor2021} and
\texttt{simplePHENOTYPES}~\citep{fernandes2020}.

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales on standard computers.
Supplementary Fig~\ref{fig:time} shows how trait simulation time scales
with number of individuals on human-like coalescent simulations
generated using~\texttt{stdpopsim}~\citep{adrion2020}.
For each ARG, we simulate a trait with 1000 causal sites
and report the mean CPU (Intel(R) Core(TM) i9-11900H)
time over 10 independent replicates.
To emphasise scalability, we also
% TODO check that the numbers align here, we may need to be 
% less specific to avoid confusing readers or having to
% explain technical details.
applied \texttt{tstrait} to the simulations of 2.7 million
French-Canadians discussed in the introduction.
% tstrait simulates phenotypes for 2.7 million individuals in the French Canadian dataset, as ancestors are denoted as individuals in the dataset.
% TODO redo this for 100 causal sites @daiki.
It took 80.69 seconds to simulate a trait with 100 causal sites.
% Is this worth the space? We should cite 1KG project also really.
Finally, to demonstrate that \texttt{tstrait} can also be applied
to ARGs inferred from real data, we simulated a trait with 100 causal
sites for an ARG estimated from 1000 Genomes project
data~\citep{kelleher2019} which has 2,504 samples
and 1,685,401 variant sites. This took 5.40 seconds.
Memory requirements for \texttt{tstrait} are modest: all of the
above experiments were performed on a laptop computer with 16GB of RAM.

@@

\begin{figure*}[t]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=\columnwidth]{figures/computational-time-code/stdpopsim-time.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds with 95\% confidence intervals plotted.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=0$. \texttt{tstrait} is used to simulate quantitative traits from ARGs simulated using \texttt{stdpopsim} with the generic piecewise constant size model in Homo Sapiens catalog. For each replicate we simulate one quantitative trait with 1000 causal sites. We have compared \texttt{tstrait} against the quantitative trait simulation framework that is based on \texttt{tskit} functionalities. The comparison codes are adapted from \citep{martin2017}, and they are modified based on functionalities offered in \texttt{tskit} version 0.5.6 to speed up the computational efficiency.}\label{fig:time}
\end{figure*}
\end{document}

>>>>

ARG-based methods can simulate genetic variation for millions of
samples and store the output very compactly
in the ``succinct tree sequence'' encoding~\citep{wong2023general}
and \texttt{tskit} library~\citep{ralph2020}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian samples~\citep{anderson2023}
requires around 550GB of storage space in gzip-compressed
VCF~\citep{danecek2011}.
At 1.36GB the original simulated ARG
(compressed using the \texttt{tszip} utility) is around 400X smaller.
Furthermore, many calculations can
be expressed efficiently in terms of the underlying
ARG~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.
Finally, outputting a simulated ARG provides access to the full
history, not just the genetic variation among the samples.

@@

\subsection{Implementation and validation}
\texttt{Tstrait} is written entirely in Python. Numerical operations
are either peformed using standard array-oriented
approaches~\citep{numpy} or accelerated
using the \texttt{numba} JIT compiler~\citep{numba}.
The \texttt{tstrait} codebase includes a suite of unit tests,
which are automatically run as part of the development process. The
output of \texttt{tstrait} has been validated against
theoretical expectations, as well as the output
of \texttt{AlphaSimR}~\citep{gaynor2021} and
\texttt{simplePHENOTYPES}~\citep{fernandes2020}.

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales on standard computers.
Supplementary Fig~\ref{fig:time} shows how trait simulation time scales
with number of individuals on human-like coalescent simulations
generated using~\texttt{stdpopsim}~\citep{adrion2020}.
To emphasise scalability, we also
% TODO check that the numbers align here, we may need to be
% less specific to avoid confusing readers or having to
% explain technical details.
applied \texttt{tstrait} to the large
French-Canadian simulations discussed in the introduction.
It took 80.69 seconds to simulate a trait with 100 causal sites
for all 2.7 million pedigree individuals.
% Is this worth the space? We should cite 1KG project also really.
Finally, to demonstrate that \texttt{tstrait} can also be applied
to ARGs inferred from real data, we simulated a trait with 100 causal
sites for an ARG estimated from 1000 Genomes project
data~\citep{kelleher2019} which has 2,504 samples
and 1,685,401 variant sites. This took 5.40 seconds.
Memory requirements for \texttt{tstrait} are modest: all of the
above experiments were performed on a laptop computer with 16GB of RAM.

@@

\begin{figure}[t]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=\columnwidth]{figures/computational-time-code/stdpopsim-time.pdf}
\caption{Time taken to simulate quantitative traits with increasing
sample size. For each sample size, we simulated an ARG under human-like
parameters using the default HomSap demographic model in \texttt{stdpopsim}.
Each point represents the mean time for 10 independent runs of \texttt{tstrait}
for a particular ARG. The times reported are the total CPU
time required to simulate a quantitative trait with 1000 causal sites,
on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM.
The trait model is a normal distribution with $\mu=0$,
$\sigma^2=1$, $h^2=0.3$, and $\alpha=0$.
}\label{fig:time}
\end{figure}
\end{document}
